Impact of Demographics and Skills of the Observer on Visual Statistical Inference

### 2013-05-30 sociology_paper.tex
brief description of experimental design

\subsection{Visual Statistical Inference} The fundamental concept of visual statistical inference is introduced by \citet{buja:2009}. In their paper, they proposed lineup protocol to discover significant pattern in the data. Later these concepts are further developed by defining necessary terminologies and validating using a head to head comparison with conventional inference by \citet{majumder:2013}. The power of the visual test is obtained and it is observed that the power is as good as conventional test while in some scenarios, the power exceeds the conventional test power.

The test statistic used in visual inference is a plot of the data. This plot is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the actual data plot in the lineup, the hypothesis is rejected. Figure \ref{fig:lineup_geno} displays a lineup of 20 plots where one of the plots is observed data. The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups. If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand if the null hypothesis is not true, observed data should be different from the rest of the plots in the lineup.

@@

This paper is organized as follows; Section \ref{sec:factor_performance} describes some of the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the design of the simulation experiments and methods to estimate the learning trend of the observer and location effect of the observed plot in a lineup. Section \ref{sec:result_socio} presents the analysis of the experimental data and the results.

@@

\subsection{Choice of Visual Test Statistic} Visual test statistics as defined in \citep{majumder:2013} serve two main purposes. One is to display a prominent pattern that should direct the human observers to select the actual plot in the lineup when the null hypothesis is not true. The another feature is that it should not show any strange pattern when null hypothesis is true, i.e., it should show similar pattern that null plots may have. Thus a visual test statistic is highly associated with the hypothesis. To achieve these purposes it is very important to decide which plot type and plot features should be adopted. As discussed in \citep{majumder:2013} following grammar of graphics developed  by \citet{wilkinson:1999} and later improved by \citet{hadley:2009} can give the task a standard control.

Some of the effective features of visual test statistics are discussed in \citep{heike:2012} including plot type, color and shape of the plots. It is also observed that a scatterplot may do a better job than a box plot when using as a visual test statistic for regression parameters \citep{majumder:2013} .  \citep{niladri:2012} presents some \textit{distance measure} to determine how a plot may be different from another.

@@

\subsection{Question that Human Observer Answers} When a observer is prepared to evaluate a lineup, there needs to be a guideline what exactly should be looked for in a lineup. The researcher knows about the hypothesis but the observer does not know the underlying details of the lineup. Thus the researcher needs to ask a question observer is to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments done by  \citet{majumder:2013}. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

@@

Given this pattern of human eye tracks, it may effect the performance of an observer depending on where the actual plot is placed in the lineup. If the actual plot is on the top left corner and the observer starts from that point it may be easier to detect the actual plot earlier in the exercise and the observer could get plenty of time to make comparison. Thus the position of the actual plot in a lineup has some impact.

\subsection{Demographics of Observer}  Age, gender and location of the observer.

\subsection{Individual Performance of the Observer} Each person is different from others in some way. For example, in a controlled experimental study \citet{zhao:2012}  notice that some people spend a lot of time to decide no matter whether the lineup is difficult or easy while some simply glance at lineups to make a decision. This influences the response of the observer and we see the subject specific variation in the power of visual test \citep{majumder:2013}.

@@

\subsection{Experiment Setup} Description of the simulation experiment giving focus on demographics and skills of experimental subjects.

%\subsection{Estimation of sample size}  How the sample size is estimated
\subsection{Data collection methods}  Description of the simulation experiment giving focus on demographics and skills of experimental subjects.
\subsection{Amazon Mechanical Turk}

Amazon Mechanical Turk \cite{turk} or MTurk is an online work place where people from around the world can perform some tasks and get paid. Usually tasks are very simple and no specialized training is required. Being a human is the main requirement. Tasks are designed for anyone to do but some tasks may require that workers satisfy some skill level depending on the recruiters need. The tasks are designed such that it does not take much time to complete. Humans are still better than computers in performing these types of tasks. The the amount of money paid for each task is very small as well.

@@

@@

@@

\subsection{Overview of the Data}

@@

\subsection{Diversity of the Experimental Subjects} Amazon Mechanical Turk \cite{turk} web site is a good source for recruiting people to evaluation lineups. It is a source of diverse workers who can give us feedback on visual inspection.  Figure \ref{fig:turker_location} shows the location from where we received data. Notice that we received data from around the world and we observed diversity of location in the collected data. We notice diversity in not only location of the observers but also their gender, age groups and education levels. Table \ref{tbl:demographics} shows that number of male and female participants are almost same.

@@

To accomplish this we fit generalized mixed effect model presented in \cite{majumder:2013} with lineups and subjects as random effect and no fixed effect. This gives the estimation of proportion correct responses  and we obtain residuals of the fitted model. The residuals give variations in proportion correct adjusted for lineup difficulties and individual skills. If there is any learning trend over different attempt $A_k$, the residuals should have that information since $A_k$ was not included as a covariate in the model. %At this point we focus on absolute residuals over different attempts to study the deviations due to attempts. An smaller absolute residual at attempt $A_k$ should indicate learning from all other attempts less than $k$.

@@

The overall proportion of correct response does not seem to vary much for different location indicating the non-existence of location effect. Since the actual data plot is same for each of the null sets of plot, the response data constitutes a multivariate response. Thus, to examine if the difference in proportion correct among the locations is statistically significant we fit a one way multi variate analysis of variance (MANOVA) model to the data.

Model \eqref{manova} enables to test if the mean vectors are similar for different locations. To fit the model we use anova function of stats package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05 and we failed to reject the hypothesis that there is no difference in location.

@@

With controlled simulation experiments, it is observed that the performance of human observers does not increase through successive attempts while evaluating multiple lineups. Even though the performance does not significantly improve, their skill in evaluating the lineup in shorter time gets improved over successive evaluations. When an observer evaluates multiple lineups, the earlier evaluations take significantly longer than the later evaluations. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. This investigation suggest that the skilled person may only do it faster.

The another simulation experiment reveals that the effect of location of actual data plot in the lineup is not significant. This is important as the visual statistical inference procedure suggest that the data plot be placed at random anywhere in the lineup.
This paper suggest that any random place in a lineup is as good as other places in the lineup.

More experiment may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference is higher with skilled observer.

>>>>

\subsection{Visual Statistical Inference} The fundamental concept of visual statistical inference is introduced by \citet{buja:2009}. In their paper, they proposed lineup protocol to discover significant pattern in the data. Later these concepts are further developed by defining necessary terminologies and validating using a head to head comparison with conventional inference by \citet{majumder:2013}. The power of the visual test is obtained and it is observed that the power is as good as that of conventional test while in some scenarios, the power exceeds the conventional test power.

The test statistic used in visual inference is a plot of the observed data. This plot, called actual data plot, is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the actual data plot in the lineup, the hypothesis is rejected. Figure \ref{fig:lineup_geno} displays a lineup of 20 plots where one of the plots is observed data. The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups. If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand if the null hypothesis is not true, observed data should be different from the rest of the plots in the lineup.

@@

This paper is organized as follows; Section \ref{sec:factor_performance} describes some of the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the design of the simulation experiments and methods to estimate the learning trend of the observer and location effect of the actual plot in a lineup. Section \ref{sec:result_socio} presents the analysis of the experimental data and the results.

@@

\subsection{Choice of Visual Test Statistic}  The visual test statistic should be highly associated with the hypothesis under investigation. To achieve this purposes it is very important to decide which plot type and plot features should be adopted.  In a linear regression setting, the visual test statistics are presented in \citep{majumder:2013} for some common hypothesis testing.  It is also observed that a scatterplot may do a better job than a box plot when using as a visual test statistic for regression parameters.  Some of the effective features of visual test statistics are discussed in \citep{heike:2012} including plot type, color and shape of the plots.  \cite{niladri:2012} presents some \textit{distance measure} to determine how a plot may be different from each other.

@@

\subsection{Question that Human Observer Answers} The researcher knows about the underlying hypothesis but the observer does not necessarily know the underlying details of the lineup. So, the researcher needs to ask a question observer is to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments done by  \citet{majumder:2013}. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

@@

Given this pattern of human eye tracks, it may effect the performance of an observer depending on where the actual plot is placed in the lineup. If the actual plot is on the top left corner and the observer starts from that point it may be easier to detect the actual plot earlier in the exercise and the observer could get plenty of time to make comparison. Thus the position of the actual plot in a lineup may have some impact.

\subsection{Demographics of Observer}  Age, gender and geographical location of the observer.

\subsection{Individual Performance of the Observer} Each person is different from others in some way. For example, in a controlled experimental study \citet{zhao:2012}  notice that some people spend a lot of time to decide no matter whether the lineup is difficult or easy while some simply glance at lineups to make a decision. This influences the response of the observer. Also subject specific variation in the power of visual test is observed in \citep{majumder:2013}.

@@

\subsection{Experiment Setup}  Multiple simulation experiments are designed to investigate the learning trend of the observer and the location effect of the actual data in the lineup. These are explained in tis section.

\paragraph{Design for Learning Trend:} Learning trend of an observer can be observed in terms of performance over successive feedback when multiple lineups are shown for evaluation. Each subject is shown a total of 10 lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups is randomized. The responses of the lineups are recorded by attempt 1 through 10. Attempt 1 means the response is for the first lineup and attempt 10 means the response is for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

\paragraph{Design for Location effect:} The experiment is set up based on the gene expression data with two groups. The groups are defined based on the presence or absence of iron sufficiency. The null hypothesis is that there is no difference between this two groups. Null plots are obtained by randomizing the group structure of the data. The actual plot is then randomly placed in five different locations. For each location, five different sets of null plots are used to produce five different lineups. This produces 25 different lineups. One such lineup is shown in Figure \ref{fig:lineup_geno}.Similar procedure is applied to obtain 25 lineups to test the interaction between Genotype and Empty Vector (EV).

Each observer is shown three lineups. First lineups shown is a randomly selected Interaction lineup, then one from randomly selected Genotype lineup. Finally a test lineup is shown to screen out unusual evaluation. The subject is not aware which one is the test plot but it is informed in the starting of the task about having a test plot. The test plot is very easy and everyone should detect the data plot. The MTurk worker is paid based on whether he or she could correctly detect the data plot. But the data on test plot are excluded from the analysis.

\subsection{Data collection methods}  The human observers who evaluate the experimental lineups are recruited through Amazon Mechanical Turk \cite{turk} or MTurk Web site.  It is an online work place where people from around the world can perform some tasks and get paid. Usually tasks are very simple and no specialized training is required. Being a human is the main requirement. Tasks are designed for anyone to do but some tasks may require that workers satisfy some skill level depending on the recruiters need. The tasks are designed such that it does not take much time to complete. Humans are still better than computers in performing these types of tasks. The the amount of money paid for each task is very small as well.

@@

We design and develop a web site which enables to display the lineups to the observers as per experimental design. The MTurk workers are redirect to the web site and the data are collected and automatically stored into a local server. Demographic informations such as age group, gender and education levels were collected. The time taken for each evaluation is computed based on the time the plot was shown and the time the feedback is received. The location of the observer is determined by the ip address of the observer.

@@

Since the actual data plot is same for each of the null sets of plot, the response data constitutes a multivariate response. Thus, to examine if the difference in proportion correct among the locations is statistically significant we fit a one way multi variate analysis of variance (MANOVA) model to the data.

@@

\subsection{Overview of the Data} Figure \ref{fig:turker_location} shows the location from where we received data. Notice that we received data from around the world and we observed diversity of location in the collected data. We notice diversity in not only location of the observers but also their gender, age groups and education levels. Table \ref{tbl:demographics} shows that number of male and female participants are almost same.

@@

@@

To accomplish this we fit generalized mixed effect model \eqref{eqn:trend} with lineups and subjects as random effect and no fixed effect. This gives the estimation of proportion correct responses  and we obtain residuals of the fitted model. The residuals give variations in proportion correct adjusted for lineup difficulties and individual skills. If there is any learning trend over different attempt $A_k$, the residuals should have that information since $A_k$ was not included as a covariate in the model. %At this point we focus on absolute residuals over different attempts to study the deviations due to attempts. An smaller absolute residual at attempt $A_k$ should indicate learning from all other attempts less than $k$.

@@

The overall proportion of correct response does not seem to vary much for different location indicating the non-existence of location effect. Model \eqref{manova} is used to test if the mean vectors are similar for different locations. To fit the model we use anova function of stats package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05 and we failed to reject the hypothesis that there is no difference in location.

@@

With controlled simulation experiments, it is observed that the performance of human observers does not increase through successive attempts while evaluating multiple lineups. But their skill in evaluating the lineup in shorter time gets improved over successive evaluations. When an observer evaluates multiple lineups, the earlier evaluations take significantly longer than the later evaluations. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. This investigation suggest that the skilled person may only do it faster.

The another simulation experiment reveals that the effect of location of actual data plot in the lineup is not significant. This is important as the visual statistical inference procedure suggests that the data plot be placed at random anywhere in the lineup. This paper suggests that any random place in a lineup is as good as other places in the lineup.

More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference is higher with skilled observer.
